---
title: "ReadMe"
author: "Will White"
date: "11/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What do these files do?

This Readme is associated with files created for the Oyster Population Sustainability project at the Guana Tolomato Matanzas National Estuarine Research Reserve (GTM NERR), led by PI Will White, co-PI David Kimbro, and postdoctoral scholar Laura Storch.

There are two main components: the R package `NERRprojectpackage` implements a population model analysis of data from oyster populations at 7 sites in GTM to assess the relative contribution to population sustainability of oysters at each site. This information is intended to be useful in planning restoration and management decisions.

The data required by `NERRprojectpackage` are oyster mortality rates and growth rates at each site. The Rmarkdown document `NERR_oyster_growth_mortality.Rmd` (also displayed in a more human-friendly version as `NERR_oyster_growth_mortality.html`) is a guide for how to analyze data to obtain those model parameters. The intent is for this to be adaptable for future data collected at GTM, or in any location.

The data files used by `NERR_oyster_growth_mortality.Rmd` are `2018_outplant.xlsx` and `2019_outplant.xlsx`. Each of these data files has a Readme tab providing metadata. To use code like that in NERR_oyster_growth_mortality.Rmd, data files should be formatted in a similar manner.

## Where to get the files

The most up-to-date version of the package and other files can be downloaded from GitHub at https://github.com/jwilsonwhite/NERR_science. 

## NERRprojectpackage

After downloading the `NERRprojectpackage` folder from the git repository, place the folder in the desired location. Or, below is code to load it directly from Github. 

If you do not have the `devtools` package, you will need to install that first:

```{r message = FALSE}
require("devtools")
library('devtools')
```

After installing devtools, from the parent working directory containing the `NERRprojectpackage` folder, in your R console type:

```{r eval=FALSE}
install("NERRprojectpackage",build_vignettes=TRUE)
# The 'build_vignettes' command ensures that the how-to vignette is also loaded
```

Alternatively, you can install the package directly from the git repository (this also requires the `devtools` package): 

```{r eval=FALSE}
install_github('NERRprojectpackage','jwilsonwhite/NERR_science/NERRprojectpackage',build_vignettes=TRUE)
```

Note that both of these steps are different from the `install.packages()` command used to install standard R packages found on CRAN. 

After installing the package, load it into your workspace as usual:

```{r}
library(NERRprojectpackage)
```

To get started, read the 'GTM' vignette that comes with the package

```{r message=FALSE}
vignette('GTM')
```


## NERR_oyster_growth_mortality

The R code in `NERR_oyster_growth_mortality.Rmd` is separate from the `NERRprojectpackage` package, and operates as standalone code. It was created using Rmarkdown. Rmarkdown allows the blending of text and code and the production of user-friendly documents (like this one). The `rmarkdown` package is not necessary to use the code, but is necessary to 'knit' updates of the code into new documents.

The code in the file works as standard R code and lists the required additional packages. The only additional requirement is that it be run in the same working directory as the two .xlsx data files.



